---
title: "MBA_Manuscript"
author: "Sabrina Grant"
date: "`r Sys.Date()`"
output: html_document
---
#Import packages and data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#installing packages
librarian::shelf("tidyverse", "vegan", "ggplot2", "dplyr", "car", "ggpubr", "minpack.lm", "googledrive", "janitor", "nlme", "RColorBrewer", "splitstackshape", "lubridate", "matrixStats", "readr", "stringr", "broom", "MASS", "caret", "grid", "reshape2", "viridis", "tibble", "ggpattern", "pairwiseAdonis")

#load data
drive_deauth()
drive_auth(scopes = "https://www.googleapis.com/auth/drive")


file <- drive_find("kelp_recovery_data.rda",
                   shared_drive = "MBA_kelp_recovery_research")
drive_download(
  as_id(file$id),
  path = "kelp_recovery_data.rda",
  overwrite = TRUE
)

load("kelp_recovery_data.rda")
ls()
```

#Data frame cleaning
```{r}
###averaging counts and % cover per transect

#helper function to get the most common value (mode)
get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

#averaging across quads
quad_transect_avgs <- quad_data %>%
  group_by(site, site_type, zone, survey_date, transect) %>%
  summarise(
    across(where(is.numeric) & !any_of("quadrat"), ~ mean(.x, na.rm = TRUE)),
    substrate = get_mode(substrate),
    .groups = "drop")

#now averaging per survey 
quad_avgs <- quad_transect_avgs %>%
  group_by(site, site_type, zone, survey_date) %>%
  #creating new column that has proportion of concealed urchins
  mutate(
    prop_purp_concealedm2 = if_else(
      purple_urchin_densitym2 > 0,
      purple_urchin_conceiledm2 / purple_urchin_densitym2,
      0),
    prop_red_concealedm2 = if_else(
      red_urchin_densitym2 > 0,
      red_urchin_conceiledm2 / red_urchin_densitym2,
      0)) %>%
  #dropping the old concealed density columns
  dplyr::select(-purple_urchin_conceiledm2, -red_urchin_conceiledm2) %>%
  #summarise by group
  summarise(
    across(where(is.numeric) & !any_of("transect"), ~mean(.x, na.rm = TRUE)),
    substrate = get_mode(substrate),
    .groups = "drop" )


#averaging kelp transects
kelp_avgs <- kelp_data %>%
  group_by(site, site_type, zone, survey_date) %>%
  summarise( 
    across(where(is.numeric) & !any_of("transect"), ~mean(.x, na.rm = TRUE)), 
    .groups = "drop") 

#combining quad and kelp data frames 
merged_data <- kelp_avgs %>%
  #keeping coordinates from kelp_avgs
  left_join(
    quad_avgs %>% dplyr::select(-latitude, -longitude),  # drop coords from quad_avgs before join
    by = c("site", "site_type", "zone", "survey_date")
  ) %>%
  mutate(year = as.numeric(format(survey_date, "%Y")))

#creating a new column that just has years for future analysis  
merged_data$year <- as.numeric(format(merged_data$survey_date, "%Y"))

#standardizing kelp counts to quad data 

#identify columns that contain "20m2"
cols_to_scale <- grep("20m2", names(merged_data), value = TRUE)

#divide those columns by 20 to get density / m2
merged_data[cols_to_scale] <- lapply(merged_data[cols_to_scale], function(x) x / 20)

#rename those columns by removing "20m2_"
names(merged_data)[names(merged_data) %in% cols_to_scale] <-
  gsub("20m2","", cols_to_scale)

#replaces NAs with zero
merged_data[sapply(merged_data, is.numeric)] <-
  lapply(merged_data[sapply(merged_data, is.numeric)],
         function(x) { x[is.na(x)] <- 0; x })  
```

#Normalizing (z-score) data to compare UPC and Swath algae
```{r}
norm_data <- merged_data %>%
  mutate(year = as.factor(year)) %>% #making year a factor instead of numeric
  mutate(site_type = as.factor(site_type)) %>%
  dplyr::select(site, site_type, zone, year, where(is.numeric)) %>%
  dplyr::select(-any_of(c("longitude", "latitude", "relief_cm", "risk_index"))) %>% #getting rid of columns I don't care about
  dplyr::mutate(across(where(is.numeric), scale)) #z-scoring data
```

#2024 data 
```{r}
#filtering for 2024 data 
year_one_data <- merged_data %>% 
  filter(year == 2024) %>%
  dplyr::select(year, site, zone, site_type, macro_stipe_density_, everything()) %>%
  mutate(site_type = as.factor(site_type))


#getting mean of numeric values across grouping categories
year_one_data %>%
  group_by(site, site_type, zone, year) %>%
  summarise(
    across(where(is.numeric), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop")
```
#2024 data (z-scored) 
```{r}
#filtering for 2024 data 
z_year_one_data <- norm_data %>% 
  filter(year == 2024) %>%
  dplyr::select(year, site, zone, site_type, macro_stipe_density_, everything())

#getting mean of numeric values across grouping categories
z_year_one_data %>%
  group_by(site, site_type, zone, year) %>%
  summarise(
    across(where(is.numeric), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop")
```
#Q1: Distinct algal communities per patch type 
####LDA on 2024 data at the depth zone level
```{r}
#run lda model
#first, i want to try it with as few variables as possible and focus mainly on the kelp species. this will tell me if macroalgae is a good indicator of patch type 
lda_model_1 <- lda(site_type ~ macro_stipe_density_ + density_ptecal + density_nerlue + density_lamset + density_eisarb + lamr + macr + macj + nerj + ptej + lsetj + cov_articulated_coralline + cov_crustose_coralline + cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + cov_dictyoneurum_spp  + cov_desmarestia_spp + cov_lam_holdfast_live + cov_mac_holdfast_live, data = year_one_data)

#lda summary
lda_model_1

#getting predictions etc etc 
pred_1 <- predict(lda_model_1)
pred_class_1 <- pred_1$class


#getting % accuracy
confusionMatrix(pred_class_1, year_one_data$site_type)
#predicts at 81% accuracy 

```
####LDA Model of 2024 Patch Types
```{r}
# Get LDA scores 
lda_scores <- as.data.frame(predict(lda_model_1)$x) 
lda_scores$site_type <- year_one_data$site_type 

# Plot LD1 vs LD2 
ggplot(lda_scores, aes(x = LD1, y = LD2, color = site_type, shape = site_type)) + geom_point(size = 3.5, alpha = 0.9) + 
  stat_ellipse(level = 0.95, size = 1, alpha = 0.6) + 
  scale_color_brewer(palette = "Set1") + 
  theme_classic(base_size = 16) + 
  labs( title = "", subtitle = "", x = paste0("LD1 (", round(lda_model$svd[1]^2 / sum(lda_model$svd^2) * 100, 1), "%)"), y = paste0("LD2 (", round(lda_model$svd[2]^2 / sum(lda_model$svd^2) * 100, 1), "%)"), color = "Site Type", shape = "Site Type" ) + 
  theme( plot.title = element_text(face = "bold", size = 20, hjust = 0.5, margin = margin(b = 10)), plot.subtitle = element_text(size = 14, hjust = 0.5, margin = margin(b = 15)), axis.title = element_text(face = "bold", size = 15), axis.text = element_text(size = 13), legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), panel.grid.minor = element_blank(), panel.grid.major = element_line(color = "white", linewidth = 0.3) )
```
####Coefficients of LDA model
```{r}
# Get coefficients into tidy format
coef_df <- as.data.frame(lda_model_1$scaling)
coef_df$variable <- rownames(coef_df)
coef_long <- pivot_longer(coef_df, cols = c(LD1, LD2), names_to = "LD", values_to = "coefficient")
```
##LDA on Z-Scored Data **using this from here on out 
```{r}
lda_model_z <- lda(site_type ~ macro_stipe_density_ + density_ptecal + density_nerlue + density_lamset + density_eisarb + density_lamstump + density_macstump + lamr + macr + macj + nerj + ptej + lsetj + cov_articulated_coralline + cov_crustose_coralline + cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + cov_dictyoneurum_spp  + cov_desmarestia_spp + cov_lam_holdfast_live + cov_mac_holdfast_live, data = z_year_one_data)

#lda summary
lda_model_z

#getting predictions etc etc 
pred_z <- predict(lda_model_z)
pred_class_z <- pred_z$class


#getting % accuracy
confusionMatrix(pred_class_z, z_year_one_data$site_type)
#predicts at 81% accuracy which is the same as non z scored
```

##Supplemental Figure 1: LDA Model 
```{r}
# Get LDA scores 
lda_scores_z <- as.data.frame(predict(lda_model_z)$x) 
lda_scores_z$site_type <- z_year_one_data$site_type 

# Plot LD1 vs LD2 
ggplot(lda_scores_z, aes(x = LD1, y = LD2, color = site_type, shape = site_type)) + geom_point(size = 3.5, alpha = 0.9) + 
  stat_ellipse(level = 0.95, size = 1, alpha = 0.6) + 
  scale_color_brewer(palette = "Set1") + 
  theme_classic(base_size = 16) + 
  labs( title = "", subtitle = "", x = paste0("LD1 (", round(lda_model_z$svd[1]^2 / sum(lda_model_z$svd^2) * 100, 1), "%)"), y = paste0("LD2 (", round(lda_model_z$svd[2]^2 / sum(lda_model_z$svd^2) * 100, 1), "%)"), color = "Site Type", shape = "Site Type" ) + 
  theme( plot.title = element_text(face = "bold", size = 20, hjust = 0.5, margin = margin(b = 10)), plot.subtitle = element_text(size = 14, hjust = 0.5, margin = margin(b = 15)), axis.title = element_text(face = "bold", size = 15), axis.text = element_text(size = 13), legend.title = element_text(face = "bold", size = 14), legend.text = element_text(size = 12), panel.grid.minor = element_blank(), panel.grid.major = element_line(color = "white", linewidth = 0.3) )
```
###Supplemental Table 1: Coefficients 
```{r}
# Get coefficients into tidy format
coef_z <- as.data.frame(lda_model_z$scaling)
coef_z$variable <- rownames(coef_z)
coef_long <- pivot_longer(coef_z, cols = c(LD1, LD2), names_to = "LD", values_to = "coefficient")
```





##Figure 2a: Density of kelp across patch types in 2024
```{r}
#mapping adult + juvenile + recruit columns to a species label
kelp_map <- tribble(
  ~species,                  ~Adults,              ~Juveniles,     ~Recruits,
  "Nereocystis luetkeana",   "density_nerlue",        "nerj",         NA_character_,
  "Pterygophora californica","density_ptecal",        "ptej",         NA_character_,
  "Laminaria setchellii",    "density_lamset",        "lsetj",        NA_character_,
  "Eisenia arborea",         "density_eisarb",        "eisj",         NA_character_,
  "Macrocystis pyrifera",    "macro_stipe_density_",  "macj",         "macr",
  "Laminariales spp.",       NA_character_,           NA_character_,  "lamr")

#building a NA-safe vector of column names to pull from year_one_data
kelp_cols <- kelp_map %>%
  dplyr::select(Adults, Juveniles, Recruits) %>%
  unlist(use.names = FALSE) %>%
  as.character() %>%
  na.omit() %>%
  unique()

kelp_lookup <- kelp_map %>%
  pivot_longer(
    cols = c(Adults, Juveniles, Recruits),
    names_to = "stage",
    values_to = "var",
    values_drop_na = TRUE)

kelp_long <- year_one_data %>%
  dplyr::select(site_type, all_of(kelp_cols)) %>%
  pivot_longer(-site_type, names_to = "var", values_to = "density") %>%
  left_join(kelp_lookup, by = "var") %>%
  filter(!is.na(species))


#getting the density of nereo adults and juvenles at each patch type in 2024
kelp_avgs <- kelp_long %>%
  mutate(
    site_type = factor(site_type, levels = c("BAR", "INCIP", "FOR")),
    stage     = factor(stage, levels = c("Adults", "Juveniles", "Recruits")),
    species   = factor(species)) %>%
  group_by(site_type, species, stage) %>%
  summarise(
    n    = sum(!is.na(density)),
    mean = mean(density, na.rm = TRUE),
    se   = sd(density, na.rm = TRUE) / sqrt(n),
    .groups = "drop")

#plotting
ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = stage)) +
  geom_col(position = position_dodge(width = 0.57), 
           width = 0.5,
           color = "black",
           linewidth = 0.35) + 
  geom_errorbar(aes(ymin = pmax(mean-se,0),ymax=mean+se), 
                position = position_dodge(width=0.57), 
                width = 0.18, 
                linewidth = 0.35)+ 
facet_wrap(~species, scales = "free_y") + 
  scale_fill_brewer(palette = "Paired", drop = FALSE) +
  ylab("Density (individuals / m²)") +
  xlab("Patch Type (2024)") +
  labs(fill = "Life Stage") +
  theme_classic(base_size = 11)


pd <- position_dodge2(width = 0.6, preserve = "single")

ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = species)) +
  geom_col(
    position = pd,
    width = 2,
    color = "black",
    linewidth = 0.35) +
  geom_errorbar(
    aes(ymin = pmax(mean - se, 0), ymax = mean + se),
    position = pd,
    width = 2,
    linewidth = 0.35) +
  facet_wrap(~ stage, scales = "free_y") +
  ylab("Density (individuals / m²)") +
  xlab("Patch Type (2024)") +
  labs(fill = "Species") +
  theme_classic(base_size = 11) +
  scale_fill_brewer(palette = "Paired", drop = TRUE) +
  theme(
    panel.spacing = grid::unit(0.6, "lines"),
    legend.position = "top",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9))

ggsave(
  "Fig_kelp_density_by_stage.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)

```

##Figure 2b: Density of kelp across patch types in 2025
```{r}

kelp_long <- year_two_data %>%
  dplyr::select(site_type, all_of(kelp_cols)) %>%
  pivot_longer(-site_type, names_to = "var", values_to = "density") %>%
  left_join(kelp_lookup, by = "var") %>%
  filter(!is.na(species))


#getting the density of nereo adults and juvenles at each patch type in 2024
kelp_avgs <- kelp_long %>%
  mutate(
    site_type = factor(site_type, levels = c("BAR", "INCIP", "FOR")),
    stage     = factor(stage, levels = c("Adults", "Juveniles", "Recruits")),
    species   = factor(species)) %>%
  group_by(site_type, species, stage) %>%
  summarise(
    n    = sum(!is.na(density)),
    mean = mean(density, na.rm = TRUE),
    se   = sd(density, na.rm = TRUE) / sqrt(n),
    .groups = "drop")

#plotting
ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = stage)) +
  geom_col(position = position_dodge(width = 0.57), 
           width = 0.5,
           color = "black",
           linewidth = 0.35) + 
  geom_errorbar(aes(ymin = pmax(mean-se,0),ymax=mean+se), 
                position = position_dodge(width=0.57), 
                width = 0.18, 
                linewidth = 0.35)+ 
facet_wrap(~species, scales = "free_y") + 
  scale_fill_brewer(palette = "Paired", drop = FALSE) +
  ylab("Density (individuals / m²)") +
  xlab("2025 Patch Type") +
  labs(fill = "Life Stage") +
  theme_classic(base_size = 11)


pd <- position_dodge2(width = 0.6, preserve = "single")

ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = species)) +
  geom_col(
    position = pd,
    width = 2,
    color = "black",
    linewidth = 0.35) +
  geom_errorbar(
    aes(ymin = pmax(mean - se, 0), ymax = mean + se),
    position = pd,
    width = 2,
    linewidth = 0.35) +
  facet_wrap(~ stage, scales = "free_y") +
  ylab("Density (individuals / m²)") +
  xlab("2025 Patch Type") +
  labs(fill = "Species") +
  theme_classic(base_size = 11) +
  scale_fill_brewer(palette = "Paired", drop = TRUE) +
  theme(
    panel.spacing = grid::unit(0.6, "lines"),
    legend.position = "top",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9))

ggsave(
  "Fig_kelp_density_by_stage.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)
```

##Figure 2c: Density of kelps averages across years 
```{r}

kelp_long <- merged_data %>%
  dplyr::select(site_type, all_of(kelp_cols)) %>%
  pivot_longer(-site_type, names_to = "var", values_to = "density") %>%
  left_join(kelp_lookup, by = "var") %>%
  filter(!is.na(species))


#getting the density of nereo adults and juvenles at each patch type in 2024
kelp_avgs <- kelp_long %>%
  mutate(
    site_type = factor(site_type, levels = c("BAR", "INCIP", "FOR")),
    stage     = factor(stage, levels = c("Adults", "Juveniles", "Recruits")),
    species   = factor(species)) %>%
  group_by(site_type, species, stage) %>%
  summarise(
    n    = sum(!is.na(density)),
    mean = mean(density, na.rm = TRUE),
    se   = sd(density, na.rm = TRUE) / sqrt(n),
    .groups = "drop")

#plotting
ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = stage)) +
  geom_col(position = position_dodge(width = 0.57), 
           width = 0.5,
           color = "black",
           linewidth = 0.35) + 
  geom_errorbar(aes(ymin = pmax(mean-se,0),ymax=mean+se), 
                position = position_dodge(width=0.57), 
                width = 0.18, 
                linewidth = 0.35)+ 
facet_wrap(~species, scales = "free_y") + 
  scale_fill_brewer(palette = "Paired", drop = FALSE) +
  ylab("Density (individuals / m²)") +
  xlab("2024 Patch Type") +
  labs(fill = "Life Stage") +
  theme_classic(base_size = 11)


pd <- position_dodge2(width = 0.6, preserve = "single")

ggplot(kelp_avgs, aes(x = site_type, y = mean, fill = species)) +
  geom_col(
    position = pd,
    width = 2,
    color = "black",
    linewidth = 0.35) +
  geom_errorbar(
    aes(ymin = pmax(mean - se, 0), ymax = mean + se),
    position = pd,
    width = 2,
    linewidth = 0.35) +
  facet_wrap(~ stage, scales = "free_y") +
  ylab("Density (individuals / m²)") +
  xlab("2024 Patch Type") +
  labs(fill = "Species") +
  theme_classic(base_size = 11) +
  scale_fill_brewer(palette = "Paired", drop = TRUE) +
  theme(
    panel.spacing = grid::unit(0.6, "lines"),
    legend.position = "top",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9))

ggsave(
  "Fig_kelp_density_by_stage.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)

```
##Figure 3a: Prop Abundance Percent cover
```{r}
#UPC species
cov_lookup <- year_one_data %>%
  dplyr::select(site, zone, site_type,
                cov_articulated_coralline,
                cov_crustose_coralline,
                cov_encrusting_red,
                cov_fleshy_red,
                cov_stephanocystis,
                cov_dictyoneurum_spp,
                cov_desmarestia_spp,
                cov_lam_holdfast_live,
                cov_mac_holdfast_live) %>%
  #creating functional groups
  dplyr::mutate(
    Corallines = cov_articulated_coralline + cov_crustose_coralline,
    Reds      = cov_encrusting_red + cov_fleshy_red,
    Browns    = cov_stephanocystis + cov_dictyoneurum_spp + cov_desmarestia_spp,
    Holdfasts = cov_lam_holdfast_live + cov_mac_holdfast_live) %>%
  #pivoting grouped categories
  pivot_longer(
    cols = c(Corallines, Reds, Browns, Holdfasts),
    names_to = "group",
    values_to = "cover")

#averages by group  
cov_avgs <- cov_lookup %>%
  group_by(site_type, group) %>%
  summarise(
    n = sum(!is.na(cover)),
    mean = mean(cover, na.rm = TRUE),
    se = sd(cover, na.rm = TRUE) / sqrt(n),
    .groups = "drop")

#getting patch type in order
cov_avgs$site_type <- factor(
  cov_avgs$site_type,
  levels = c("BAR", "INCIP", "FOR"))

#plotting
ggplot(cov_avgs, aes(x = site_type, y = mean,
           fill = site_type,
           pattern = site_type)) +
geom_col_pattern(
    color = "black",
    pattern_fill = "grey70",
    pattern_density = 0.4,
    pattern_spacing = 0.07) +
geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2) +
facet_wrap(~ group, scales = "free_y") +
scale_fill_manual(values = c(
    "BAR"   = "white",
    "INCIP" = "white",
    "FOR"   = "grey")) +
  scale_pattern_manual(values = c(
    "BAR"   = "none",
    "INCIP" = "stripe",
    "FOR"   = "none")) +
  labs(x = "Patch type (2024)",
       y = "Mean percent cover (± SE)") +
  theme_classic() +
  theme(legend.position = "none")

ggsave(
  "avg_percent_cov.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)


#plotting
ggplot(cov_avgs, aes(x = group, y = mean,
           fill = group,
           pattern = group)) +
geom_col_pattern(
    color = "black",
    pattern_fill = "grey70",
    pattern_density = 0.4,
    pattern_spacing = 0.05) +
geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2) +
facet_wrap(~ site_type) +
scale_fill_manual(values = c(
    "Corallines"   = "white",
    "Browns" = "white",
    "Holdfasts"   = "grey", 
    "Reds" = "darkgrey")) +
  scale_pattern_manual(values = c(
     "Corallines"   = "stripe",
    "Browns" = "none",
    "Holdfasts"   = "none", 
    "Reds" = "none")) +
  labs(x = "Patch type (2024)",
       y = "Mean percent cover (± SE)") +
  theme_classic() +
  theme(legend.position = "none", 
    axis.text.x = element_text(angle = 30, hjust = 1, face = "bold", size = 11))

ggplot(cov_avgs, aes(x = group, y = mean,
           fill = group,
           pattern = group)) +
geom_col_pattern(
    color = "black",
    pattern_fill = "grey70",
    pattern_density = 0.4,
    pattern_spacing = 0.05) +
geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
                width = 0.2) +
facet_wrap(~ site_type) +
scale_fill_manual(values = c(
    "Corallines"   = "white",
    "Browns" = "white",
    "Holdfasts"   = "grey", 
    "Reds" = "darkgrey")) +
  scale_pattern_manual(values = c(
     "Corallines"   = "stripe",
    "Browns" = "none",
    "Holdfasts"   = "none", 
    "Reds" = "none")) +
  labs(x = "Patch type (2024)",
       y = "Mean percent cover (± SE)") +
  theme_classic() +
  theme(legend.position = "none", 
    axis.text.x = element_text(angle = 30, hjust = 1, face = "bold", size = 11))

#averages by species
cov_species_avgs <- cov_lookup %>%
  pivot_longer(
    cols = starts_with("cov_"),
    names_to = "species",
    values_to = "species_cover") %>%
  group_by(site_type, species) %>%
  summarise(
    n = sum(!is.na(species_cover)),
    mean = mean(species_cover, na.rm = TRUE),
    se = sd(species_cover, na.rm = TRUE) / sqrt(n),
    .groups = "drop")

sp_key <- tibble::tribble(
  ~species,                         ~code,   ~fg,
  "cov_articulated_coralline",       "ACA",    "Coralline Algae",
  "cov_crustose_coralline",          "CCA",    "Coralline Algae",
  "cov_encrusting_red",              "ER",      "Red Algae", 
  "cov_fleshy_red",                  "FR",       "Red Algae", 
  "cov_stephanocystis",              "STEPH.",    "Brown Algae", 
  "cov_dictyoneurum_spp",           "DICTYO. SPP.",    "Brown Algae",
   "cov_desmarestia_spp",            "DESMAR. SPP.", "Brown Algae", 
  "cov_lam_holdfast_live",           "LAM HF", "Kelp Holdfasts",
  "cov_mac_holdfast_live",           "MAC HF",   "Kelp Holdfasts")

fg_pal <- c(
  "Coralline Algae" = "coral1",
  "Red Algae" = "firebrick",
  "Brown Algae" = "orange4", 
  "Kelp Holdfasts" = "green4"
)

code_order <- c(
  "ACA", "CCA",        # corallines
  "ER", "FR",          # red algae
  "STEPH.", "DICTYO. SPP.", "DESMAR. SPP.",   # brown algae
  "LAM HF", "MAC HF"   # holdfasts
)

cov_plot <- cov_species_avgs %>%
  left_join(sp_key, by = "species") %>%
  mutate(code = ifelse(is.na(code), species, code)) %>% # fallback if missing 
mutate(code = factor(code, levels = code_order))

ggplot(cov_plot, aes(x = code, y = mean, fill = fg)) +
  geom_col(color="black", position = "stack") +
 # geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  facet_wrap(~ site_type) +
scale_fill_manual(values = fg_pal, drop = FALSE)+
  labs(x = "Patch type (2024)",
       y = "Mean percent cover (± SE)",
       fill = "Functional group") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 9))

#ggsave(
  "avg_percent_cov.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)



```


#Q2: Demonstrated pattern of algal dominance in kelp recovery
##LDA model on 2025 data 
```{r}
#filtering for 2025 data 
z_year_two_data <- norm_data %>% 
  filter(year == 2025) %>%
  dplyr::select(year, site, zone, site_type, macro_stipe_density_, everything())

#getting mean of numeric values across grouping categories
z_year_two_data %>%
  group_by(site, site_type, zone, year) %>%
  summarise(
    across(where(is.numeric), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop")

#predicting 2025 patch types based on 2024 LDA model
z_year_two_data <- z_year_two_data %>%
  mutate(site_type_pred = predict(lda_model_z, newdata = .)$class) %>%
  dplyr::select(year, site, zone, site_type, site_type_pred, everything())


#setting up the 2025 df with the predictions column 
z_year_one_data <- z_year_one_data %>%
  mutate(site_type_pred = predict(lda_model_z, newdata = .)$class) %>%
  dplyr::select(year, site, zone, site_type, site_type_pred, everything())


##REJOINING 2024 AND 2025 DFS
mergedz <- z_year_one_data %>%
  dplyr::select(site, zone, site_type, site_type_pred_2024 = site_type_pred) %>%
  left_join(z_year_two_data %>%
    dplyr::select(site, zone, site_type, site_type_pred_2025 = site_type_pred),
    by = c("site", "zone", "site_type"))

```
##Supplemental Figure 1: Heatmap of patch-type transitions from 2024 to 2025 (on z-scored data)
```{r}
# Melt the table
cross_tab_melt <- reshape2::melt(cross_tab)

# changing the order of the variables for my plot
desired_order_2024 <- c("BAR", "INCIP", "FOR") 
desired_order_2025 <- c("BAR", "INCIP", "FOR")  

# Convert to factors with specified levels
cross_tab_melt$Var1 <- factor(cross_tab_melt$Var1, levels = desired_order_2024)
cross_tab_melt$Var2 <- factor(cross_tab_melt$Var2, levels = desired_order_2025)

#map colors
my_colors <- brewer.pal(11, "Blues")

# Sleek publication-ready heatmap
ggplot(cross_tab_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "grey90", linewidth = 0.3) +
  geom_text(aes(label = value), 
            color = ifelse(cross_tab_melt$value > max(cross_tab_melt$value)/2, "white", "black"), 
            size = 5, fontface = "bold") +
 scale_fill_gradientn(colors = my_colors) +
  labs(
    x = "2024",
    y = "2025",
    title = "Transition of Predicted Patch Types from 2024 to 2025",
    fill = "Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 13),
    axis.text.y = element_text(face = "bold", size = 13),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 18, margin = margin(b = 10)),
    legend.title = element_text(face = "bold", size = 13),
    legend.text = element_text(size = 12),
    panel.grid = element_blank()
  )
```

#Patch transition df of normalized data
```{r}
predictions_df <- mergedz %>%
  left_join(norm_data, by = c("site", "zone", "site_type")) %>%
  dplyr::select(c(site,zone,site_type_pred_2024,site_type_pred_2025, macro_stipe_density_, density_ptecal, density_nerlue, density_lamset, density_eisarb, density_lamstump, density_macstump, lamr , macr , macj , nerj , ptej , lsetj , cov_articulated_coralline , cov_crustose_coralline , cov_encrusting_red , cov_fleshy_red , cov_stephanocystis , cov_dictyoneurum_spp  , cov_desmarestia_spp , cov_lam_holdfast_live ,cov_mac_holdfast_live,year))

patch_transitions <- predictions_df %>%
  mutate(
    # assign the correct prediction for each year
    predicted_patch_type = case_when(
      year == 2024 ~ as.character(site_type_pred_2024),
      year == 2025 ~ as.character(site_type_pred_2025),
      TRUE ~ NA_character_
    ),
    # create the transition column
    patch_transition = paste(site_type_pred_2024, "→", site_type_pred_2025)
  ) %>%
  # drop the old wide prediction columns, keep everything else (including species)
  dplyr::select(site, zone, year, predicted_patch_type, patch_transition, everything(), -site_type_pred_2024, -site_type_pred_2025)


#creating groupings that I want for nmds
patch_groupings <- patch_transitions %>%
  mutate(patches_groups = case_when(
    year == 2024 & predicted_patch_type == "BAR" ~ "2024 Barrens",
    year == 2024 & predicted_patch_type == "INCIP" ~ "2024 Incipients",
    year == 2024 & predicted_patch_type == "FOR" ~ "2024 Forests", 
    year == 2025 & patch_transition %in% c("BAR → INCIP") ~ "2025 Recovered Incipient Patches",
    year == 2025 & patch_transition %in% c("BAR → FOR", "INCIP → FOR") ~ "2025 Recovered Forest Patches",
    year == 2025 & patch_transition %in% c("BAR → BAR") ~ "2025 Persistant Barren Patches",
    year == 2025 & patch_transition %in% c("FOR → FOR") ~ "2025 Persistant Forest Patches",
    year == 2025 & patch_transition %in% c("INCIP → INCIP") ~ "2025 Persistant Incipient Patches",
    year == 2025 & patch_transition %in% c("INCIP → BAR", "FOR → BAR") ~ "2025 Regressed Barren Patches",
    year == 2025 & patch_transition %in% c("FOR → INCIP") ~ "2025 Regressed Incipient Patches",
    TRUE ~ NA_character_))
```

#Patch transition df of raw data
```{r}
raw_predictions_df <- mergedz %>%
  left_join(merged_data, by = c("site", "zone", "site_type")) %>%
  dplyr::select(c(site,zone,site_type_pred_2024,site_type_pred_2025, macro_stipe_density_, density_ptecal, density_nerlue, density_lamset, density_eisarb, density_lamstump, density_macstump, lamr , macr , macj , nerj , ptej , lsetj , cov_articulated_coralline , cov_crustose_coralline , cov_encrusting_red , cov_fleshy_red , cov_stephanocystis , cov_dictyoneurum_spp  , cov_desmarestia_spp , cov_lam_holdfast_live ,cov_mac_holdfast_live,year))

raw_patch_transitions <- raw_predictions_df %>%
  mutate(
    # assign the correct prediction for each year
    predicted_patch_type = case_when(
      year == 2024 ~ as.character(site_type_pred_2024),
      year == 2025 ~ as.character(site_type_pred_2025),
      TRUE ~ NA_character_
    ),
    # create the transition column
    patch_transition = paste(site_type_pred_2024, "→", site_type_pred_2025)
  ) %>%
  # drop the old wide prediction columns, keep everything else (including species)
  dplyr::select(site, zone, year, predicted_patch_type, patch_transition, everything(), -site_type_pred_2024, -site_type_pred_2025)


#creating groupings that I want for nmds
raw_patch_groupings <- raw_patch_transitions %>%
  mutate(raw_patches_groups = case_when(
    year == 2024 & predicted_patch_type == "BAR" ~ "2024 Barrens",
    year == 2024 & predicted_patch_type == "INCIP" ~ "2024 Incipients",
    year == 2024 & predicted_patch_type == "FOR" ~ "2024 Forests", 
    year == 2025 & patch_transition %in% c("BAR → INCIP") ~ "2025 Recovered Incipient Patches",
    year == 2025 & patch_transition %in% c("BAR → FOR", "INCIP → FOR") ~ "2025 Recovered Forest Patches",
    year == 2025 & patch_transition %in% c("BAR → BAR") ~ "2025 Persistant Barren Patches",
    year == 2025 & patch_transition %in% c("FOR → FOR") ~ "2025 Persistant Forest Patches",
    year == 2025 & patch_transition %in% c("INCIP → INCIP") ~ "2025 Persistant Incipient Patches",
    year == 2025 & patch_transition %in% c("INCIP → BAR", "FOR → BAR") ~ "2025 Regressed Barren Patches",
    year == 2025 & patch_transition %in% c("FOR → INCIP") ~ "2025 Regressed Incipient Patches",
    TRUE ~ NA_character_))
```


#nMDS of patch transitions from normalized data 
##of entire patch mosaic
```{r}

nmds_groups <- patch_groupings %>%
  dplyr::select(c(site, zone, year, patches_groups, macro_stipe_density_:cov_mac_holdfast_live)) %>%
  filter(!is.na(patches_groups)) %>% #just keeping rows with my patch transitions 
  droplevels()

#getting community matrix and metadata matrix 
meta_data <- nmds_groups %>%
  dplyr::select(c(site, zone, year, patches_groups))

comm_data <- nmds_groups %>%
  dplyr::select(-c(site, zone, year, patches_groups))
  
#euclidean distance used because zscored data has lots of negatives so the hellinger/ bray curtis did not work 
set.seed(1)
nmds <- metaMDS(comm_data, distance = "euclidean", k = 2, trymax = 999, trace = FALSE)

#centroids plot (centroids only)
scores_df <- as.data.frame(scores(nmds, display = "sites"))
scores_df$group <- factor(meta_data$patches_groups)

centroids <- scores_df %>%
  group_by(group) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    n = n(),
    .groups = "drop"
  )

ggplot(centroids, aes(NMDS1, NMDS2, label = group)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel() +
  theme_classic() +
  labs(
    title = "NMDS centroids by patch group",
    subtitle = paste0("Euclidean distance on z-scored data; Stress = ", round(nmds$stress, 3))
  )

#PERMANOVA on the same distance used by NMDS
dist_euc <- vegdist(comm_data, method = "euclidean")

perm <- adonis2(dist_euc ~ patches_groups, data = meta_data, permutations = 999)
perm

#dispersion check (report alongside PERMANOVA)
disp <- betadisper(dist_euc, meta_data$patches_groups)
anova(disp)
permutest(disp, permutations = 999)
```
##of just recovery transitions
```{r}
recovery_groups <- patch_groupings %>%
  filter(patches_groups %in% c(
    "2024 Barrens",
    "2024 Incipients",
    "2024 Forests",
    "2025 Recovered Forest Patches",
    "2025 Recovered Incipient Patches"
  ))

nmds_groups <- recovery_groups %>%
  dplyr::select(c(site, zone, year, patches_groups, macro_stipe_density_:cov_mac_holdfast_live)) %>%
  filter(!is.na(patches_groups)) %>% #just keeping rows with my patch transitions 
  droplevels()

#getting community matrix and metadata matrix 
meta_data <- nmds_groups %>%
  dplyr::select(c(site, zone, year, patches_groups))

comm_data <- nmds_groups %>%
  dplyr::select(-c(site, zone, year, patches_groups))
  
#euclidean distance used because zscored data has lots of negatives so the hellinger/ bray curtis did not work 
set.seed(1)
nmds <- metaMDS(comm_data, distance = "euclidean", k = 2, trymax = 999, trace = FALSE)

#centroids plot (centroids only)
scores_df <- as.data.frame(scores(nmds, display = "sites"))
scores_df$group <- factor(meta_data$patches_groups)

centroids <- scores_df %>%
  group_by(group) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    n = n(),
    .groups = "drop"
  )

ggplot(centroids, aes(NMDS1, NMDS2, label = group)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel() +
  theme_classic() +
  labs(
    title = "NMDS centroids by recovering patch group",
    subtitle = paste0("Euclidean distance on z-scored data; Stress = ", round(nmds$stress, 3))
  )

#PERMANOVA on the same distance used by NMDS
dist_euc <- vegdist(comm_data, method = "euclidean")

perm <- adonis2(dist_euc ~ patches_groups, data = meta_data, permutations = 999)
perm

#post hoc tests
#bar -> incip
keep <- meta_data$patches_groups %in% c(
  "2024 Barrens",
  "2025 Recovered Incipient Patches"
)

sub_meta <- meta_data[keep, , drop = FALSE]
sub_meta$patches_groups <- droplevels(as.factor(sub_meta$patches_groups))

sub_dist <- as.dist(as.matrix(dist_euc)[keep, keep])

# run PERMANOVA
pair_test <- adonis2(sub_dist ~ patches_groups,
                     data = sub_meta,
                     permutations = 999)
pair_test

#bar -> for
keep <- meta_data$patches_groups %in% c(
  "2024 Barrens",
  "2025 Recovered Forest Patches"
)

sub_meta <- meta_data[keep, , drop = FALSE]
sub_meta$patches_groups <- droplevels(as.factor(sub_meta$patches_groups))

sub_dist <- as.dist(as.matrix(dist_euc)[keep, keep])

# run PERMANOVA
pair_test <- adonis2(sub_dist ~ patches_groups,
                     data = sub_meta,
                     permutations = 999)
pair_test

#incip -> for
keep <- meta_data$patches_groups %in% c(
  "2024 Incipients",
  "2025 Recovered Forest Patches"
)

sub_meta <- meta_data[keep, , drop = FALSE]
sub_meta$patches_groups <- droplevels(as.factor(sub_meta$patches_groups))

sub_dist <- as.dist(as.matrix(dist_euc)[keep, keep])

# run PERMANOVA
pair_test <- adonis2(sub_dist ~ patches_groups,
                     data = sub_meta,
                     permutations = 999)
pair_test



#dispersion check (report alongside PERMANOVA)
disp <- betadisper(dist_euc, meta_data$patches_groups)
anova(disp)
permutest(disp, permutations = 999)



# --- NMDS scores (sites) ---
scores_df <- as.data.frame(scores(nmds, display = "sites")) %>%
  mutate(
    group = factor(meta_data$patches_groups),
    zone  = factor(meta_data$zone))

# --- Centroids ---
centroids <- scores_df %>%
  group_by(group) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    n = n(),
    .groups = "drop")

# --- Define recovery trajectories (edit if your naming differs) ---
traj <- tibble(
  from = c(
    "2024 Barrens",
    "2024 Barrens",
    "2024 Incipients"
  ),
  to = c(
    "2025 Recovered Incipient Patches",
    "2025 Recovered Forest Patches",
    "2025 Recovered Forest Patches"
  )
) %>%
  left_join(centroids, by = c("from" = "group")) %>%
  rename(x = NMDS1, y = NMDS2) %>%
  left_join(centroids, by = c("to" = "group")) %>%
  rename(xend = NMDS1, yend = NMDS2)


traj$type <- c("Barren to Incipient", "Barren to Forest", "Incipient to Forest")

# --- Plot ---
ggplot() +
  # faint points for context (comment out if you want centroids only)
  geom_point(
    data = scores_df,
    aes(NMDS1, NMDS2, shape = zone),
    alpha = 0.25, size = 2, color = "grey35"
  ) +

  # arrows showing trajectories
  geom_segment(
    data = traj,
    aes(x = x, y = y, xend = xend, yend = yend, color = type),
    arrow = arrow(length = unit(4, "mm")),
    linewidth = 2) +

  # centroids
  geom_point(
    data = centroids,
    aes(NMDS1, NMDS2),
    size = 4.2, shape = 21, fill = "white", color = "black", stroke = 1
  ) +

  # centroid labels
  ggrepel::geom_text_repel(
    data = centroids,
    aes(NMDS1, NMDS2, label = group),
    size = 3.6,
    box.padding = 0.6,
    point.padding = 0.3,
    min.segment.length = 0
  ) +

  coord_equal() +
  coord_cartesian(xlim = c(-1, 1), ylim = c(-1, 1)) +

  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  labs(
    x = "NMDS1",
    y = "NMDS2",
    shape = "Zone")


ggsave(
  "transition_nmds.png",
  width = 10,
  height = 7,
  units = "in",
  dpi = 600
)

```

##PCA of just 2024 (Q1)
```{r}
groups_2024 <- patch_groupings %>%
   filter(patches_groups %in% c(
    "2024 Barrens",
    "2024 Incipients",
    "2024 Forests") )

year_1_groups <- groups_2024 %>%
  dplyr::select(c(site, zone, patches_groups, macro_stipe_density_:cov_mac_holdfast_live)) %>%
  filter(!is.na(patches_groups)) %>% #just keeping rows with my patch transitions 
  droplevels()

#getting community matrix and metadata matrix 
meta_data <- year_1_groups %>%
  dplyr::select(c(site, zone, patches_groups))

comm_data <- year_1_groups %>%
  dplyr::select(-c(site, zone, patches_groups))
  
pca_2024 <- prcomp(comm_data, scale. = FALSE)
summary(pca_2024)
biplot(pca_2024)
  

scores <- as.data.frame(pca_2024$x)
scores$patch <- meta_data$patches_groups

ggplot(scores, aes(PC1, PC2, color = patch)) +
  geom_point(size = 3) +
  theme_classic()


set.seed(1)
adonis_result <- adonis2(
  comm_data ~ patches_groups,
  data = meta_data,
  method = "euclidean"
)

adonis_result


disp <- betadisper(dist(comm_data), meta_data$patches_groups)
anova(disp)
```

```{r}
library(ggplot2)
library(dplyr)

# variance explained
var_explained <- summary(pca_2024)$importance[2, 1:2] * 100

scores <- as.data.frame(pca_2024$x)
scores$patch <- meta_data$patches_groups

ggplot(scores, aes(PC1, PC2, color = patch)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(fill = patch),
               geom = "polygon",
               alpha = 0.2,
               color = NA) +
  stat_ellipse(level = 0.95, linewidth = 1) +
 scale_color_brewer(palette = "Set1") + 
  labs(
    x = paste0("PC1 (", round(var_explained[1],1), "%)"),
    y = paste0("PC2 (", round(var_explained[2],1), "%)"),
    color = "Patch Type",
    fill = "Patch Type"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    legend.position = "right"
  )


ggsave(
  "pca_2024.png",
  width = 10,
  height = 7,
  units = "in",
  dpi = 600
)
```

##PCA of both years (Q1)
```{r}

#getting community matrix and metadata matrix 
meta_data <- patch_groupings %>%
  dplyr::select(c(site, zone, year, predicted_patch_type))

species_mat <- patch_groupings %>%
  dplyr::select(-c(site:patch_transition, patches_groups))
  
pca <- prcomp(species_mat, scale. = FALSE)
summary(pca)
biplot(pca)
  
scores_sites <- as.data.frame(pca$x)

plot_data <- bind_cols(meta_data, scores_sites)

ggplot(plot_data,
       aes(PC1, PC2,
           color = predicted_patch_type,
           shape = year)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic()

loadings <- as.data.frame(pca$rotation)
loadings %>%
  arrange(desc(abs(PC1))) %>%
  head(10)

set.seed(1)
adonis2(species_mat ~ year * predicted_patch_type,
        data = meta_data)


scores <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  patch = meta_data$predicted_patch_type,
  year  = meta_data$year
)

var_explained <- (pca$sdev^2 / sum(pca$sdev^2)) * 100

centroids <- scores %>%
  group_by(patch, year) %>%
  summarise(
    PC1 = mean(PC1),
    PC2 = mean(PC2),
    .groups = "drop"
  )

ggplot(scores, aes(PC1, PC2, color = patch)) +

  ## sample points
  geom_point(aes(shape = year),
             size = 3,
             alpha = 0.7) +

  ## filled ellipses (community spread)
  stat_ellipse(aes(fill = patch),
               geom = "polygon",
               alpha = 0.12,
               color = NA,
               level = 0.95) +

  ## ellipse outlines
  stat_ellipse(linewidth = 1,
               level = 0.95,
               show.legend = FALSE) +

  ## centroid points
  geom_point(data = centroids,
             size = 5,
             stroke = 1.2,
             shape = 21,
             fill = "white") +

  ## temporal trajectories
  geom_path(data = centroids,
            aes(group = patch),
            linewidth = 1.2,
            arrow = arrow(length = unit(0.15,"cm"))) +

  ## colorblind-safe palette
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +

  labs(
    x = paste0("PC1 (", round(var_explained[1],1), "%)"),
    y = paste0("PC2 (", round(var_explained[2],1), "%)"),
    color = "Patch Type",
    fill  = "Patch Type",
    shape = "Year"
  ) +

  coord_equal() +   # ⭐ VERY IMPORTANT

  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(face = "bold"),
    legend.position = "right",
    legend.box = "vertical"
  )

ggsave(
  "pca_2024.png",
  width = 10,
  height = 7,
  units = "in",
  dpi = 600
)
```

#SIMPER on raw data
```{r}
raw_groups <- raw_patch_groupings %>%
  dplyr::select(c(site, zone, year, raw_patches_groups, macro_stipe_density_:cov_mac_holdfast_live)) %>%
  filter(!is.na(raw_patches_groups)) %>% #just keeping rows with my patch transitions 
  droplevels()

#just percent cover species 
raw_cover <- raw_groups %>%
  dplyr::select(-c(macro_stipe_density_:lsetj))
cover_meta <- raw_cover %>%
  dplyr::select(c(site, zone, year, raw_patches_groups))
cover_comm<- raw_cover %>%
  dplyr::select(-c(site, zone, year, raw_patches_groups))

sim_cover <- simper(cover_comm, cover_meta$raw_patches_groups, permutations = 999)

summary(sim_cover)


#just density species 
raw_density <- raw_groups %>%
  dplyr::select(c(site:lsetj))

  
den_meta <- raw_density %>%
  dplyr::select(c(site, zone, year, raw_patches_groups))
den_comm<- raw_density %>%
  dplyr::select(-c(site, zone, year, raw_patches_groups))
den_comm_eps <- den_comm + 1e-6 #adding in a tiny value to accoutn for zero inflation, dont want to get rid of zeros 
d <- vegdist(den_comm_eps, method="bray")
sim_den <- simper(den_comm_eps, den_meta$raw_patches_groups, permutations = 999)

summary(sim_den)
  



```


#random forest model 
```{r}
library(nnet)
library(randomForest)
model_state <- multinom(predicted_patch_type ~ ., data = patch_transitions)

species_2024 <- z_year_one_data %>%
  dplyr::select(c(year, 
                site,
                zone,
                site_type_pred,
                cov_articulated_coralline,
                cov_crustose_coralline,
                cov_encrusting_red,
                cov_fleshy_red,
                cov_stephanocystis,
                cov_dictyoneurum_spp,
                cov_desmarestia_spp,
                cov_lam_holdfast_live,
                cov_mac_holdfast_live, macro_stipe_density_,
                density_ptecal, 
                density_nerlue,
                density_lamset,
                density_eisarb,
                lamr,
                macr,
                macj,
                nerj,
                ptej,
                lsetj))
                


#just on 2024 data
rf_2024 <- randomForest(as.factor(site_type_pred) ~ . -year,
  data = species_2024,
  importance = TRUE)
importance(rf_2024)

rf_state <- randomForest(
  as.factor(predicted_patch_type) ~ . - patch_transition,
  data = patch_transitions,
  importance = TRUE)
importance(rf_state)

```



#Q3: Diversity and Evenness
```{r}
print(raw_patch_groupings)

for_for <- raw_patch_groupings %>%
  dplyr::select(c(site, zone, raw_patches_groups,-predicted_patch_type,-patch_transition,macro_stipe_density_:raw_patches_groups)) %>%
  filter(raw_patches_groups %in% c(
    "2025 Persistant Forest Patches",
    "2025 Recovered Forest Patches"))


```

