---
title: "SG_analyses"
author: "Sabrina Grant"
date: "`r Sys.Date()`"
output: html_document
---

#packages
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(stats)
library(dplyr)
library(janitor)
library(nlme)
library(matrixStats)
library(RColorBrewer)
library(egg)
library(vegan)
library(librarian)
library(splitstackshape)
library(lubridate)
library(piecewiseSEM)
library(matrixStats)
library(indicspecies)
library(ggh4x)
library(Rcmdr)
library(tcltk)
library(BiodiversityR)
library(ggforce)
library(concaveman)
library(ggrepel)
library(ggsci)
library(pvclust)
library(googlesheets4)
library(stringr)
library(readr)

```

```{r}
rm(list=ls())
require(tidyverse)  
require(here)

load(here::here("output","processed", "cleaned_survey_data.Rda")) 
```


#Fixing up data frames
```{r}
###averaging counts and % cover per transect

# Helper function to get the most common value (mode)
get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

quad_transect_avgs <- quad_data %>%
  group_by(site, site_type, zone, survey_date, transect) %>%
  summarise(
    across(where(is.numeric) & !any_of("quadrat"), ~ mean(.x, na.rm = TRUE)),
    substrate = get_mode(substrate),
    .groups = "drop"
  )


# View result
print(quad_transect_avgs)

#now averaging per survey 
quad_avgs <- quad_transect_avgs %>%
  group_by(site, site_type, zone, survey_date) %>%
  summarise( 
    across(where(is.numeric) & !any_of("transect"), ~mean(.x, na.rm = TRUE)), 
    substrate = get_mode(substrate), 
    .groups = "drop")

# View result
print(quad_avgs)

#now making a new column to represent urchin behavior i.e. how many urchins are actively grazing 

quad_avgs1 <- quad_avgs %>%
  mutate( purple_urchin_active_foraging_densitym2 = purple_urchin_densitym2 - purple_urchin_conceiledm2,
    red_urchin_active_foraging_densitym2 = red_urchin_densitym2 - red_urchin_conceiledm2
  ) %>%
  mutate(total_urchin_densitym2 = purple_urchin_densitym2 + red_urchin_densitym2, 
         total_urchin_foraging_densitym2 = purple_urchin_active_foraging_densitym2 + red_urchin_active_foraging_densitym2)
  #separate purps and reds, start with just purps

#averaging transects
kelp_avgs <- kelp_data %>%
  group_by(site, site_type, zone, survey_date) %>%
  summarise( 
    across(where(is.numeric) & !any_of("transect"), ~mean(.x, na.rm = TRUE)), 
    .groups = "drop") 

macro_avgs <- macro_data %>%
  group_by(site, site_type, zone, survey_date) %>%
  summarise( 
    across(where(is.numeric) & !any_of("transect"), ~mean(.x, na.rm = TRUE)), 
    .groups = "drop")

view(macro_avgs)
```



#Merging Quad/Kelp/Macro Dfs

```{r}
merged_data <- kelp_avgs %>%
  mutate(site = if_else(site == "RECO_04", "REC_04", site)) %>%
  filter(!(site == "REC_05" & site_type == "INCIP" & zone == "Shallow" & survey_date == as.Date("2025-06-09"))) %>%
  left_join(quad_avgs1, by = c("site", "site_type", "zone", "survey_date")) %>%
  left_join(macro_avgs, by = c("site","site_type", "zone", "survey_date"))

#creating a new column thta just has years for future analysis  
merged_data$year <- as.numeric(format(merged_data$survey_date, "%Y"))
  

print(merged_data)
```

#2024 data 
```{r}
#normalizing 2024 data 


year_one_data <- merged_data %>% 
  filter(year == 2024) %>%
  filter() %>%
  dplyr::select(year, site, zone, site_type, macro_stipe_density_20m2, everything())

year_one_data[sapply(year_one_data, is.numeric)] <-
  lapply(year_one_data[sapply(year_one_data, is.numeric)],
         function(x) { x[is.na(x)] <- 0; x })

year_one_data %>%
  group_by(site, site_type, zone, year) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")
  

view(year_one_data)

```


##LDA 
```{r}
library(MASS)

year_one_data$site_type <- as.factor(year_one_data$site_type)
  
# Identify columns that contain "20m2"
cols_to_scale <- grep("20m2", names(year_one_data), value = TRUE)

# Divide those columns by 20
year_one_data[cols_to_scale] <- lapply(year_one_data[cols_to_scale], function(x) x / 20)

# 3. Rename those columns by removing "20m2_"
names(year_one_data)[names(year_one_data) %in% cols_to_scale] <-
  gsub("20m2","", cols_to_scale)

lda_model <- lda(site_type ~ macro_stipe_density_ + densitymacstump + densityptecal + densitynerlue + densitylamset + densityeisarb + purple_urchin_densitym2 + purple_urchin_conceiledm2 + lamr_densitym2 + macr_densitym2 + macj_densitym2 + nerj_densitym2 + ptej_densitym2 + lsetj_densitym2  + tegula_densitym2 + cov_articulated_coralline + cov_crustose_coralline + cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + cov_dictyoneurum_spp + cov_dictyota_dictyopteris + cov_desmarestia_spp, data = year_one_data)


#lda figure
plot(lda_model)

#lda summary
lda_model


#getting predictions etc etc 
pred <- predict(lda_model)

pred_class <- pred$class

table(Actual = year_one_data$site_type, Predicted = pred_class)

results <- year_one_data %>%
  mutate(predicted = pred_class, 
         match = site_type == predicted)

view(results)

mismatches <- results %>% filter(!match)

view(mismatches)

results <- results %>%
  dplyr::select(year, site, zone, site_type, predicted, match, everything())

results 

pred <- predict(lda_model)
table(Predicted = pred$class, Actual = year_one_data$site_type)
mean(pred$class == year_one_data$site_type)

```
###testing the stats of the lda 
```{r}
#this is from chatgpt, not sure if necessary or not 

# ---- SETTINGS ----
set.seed(123)       # For reproducibility
n_perm <- 999       # Number of permutations for significance testing

# ---- LDA ----
lda_model <- lda(site_type ~ macro_stipe_density_ + densitymacstump + densityptecal + 
                   densitynerlue + densitylamset + densityeisarb + purple_urchin_densitym2 + 
                   purple_urchin_conceiledm2 + lamr_densitym2 + macr_densitym2 + 
                   macj_densitym2 + nerj_densitym2 + ptej_densitym2 + lsetj_densitym2 + 
                   tegula_densitym2 + cov_articulated_coralline + cov_crustose_coralline + 
                   cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + 
                   cov_dictyoneurum_spp + cov_dictyota_dictyopteris + cov_desmarestia_spp,
                 data = year_one_data)

cat("\n---- LDA Summary ----\n")
print(lda_model)

# ---- CROSS-VALIDATION ----
lda_cv <- lda(site_type ~ macro_stipe_density_ + densitymacstump + densityptecal + 
                densitynerlue + densitylamset + densityeisarb + purple_urchin_densitym2 + 
                purple_urchin_conceiledm2 + lamr_densitym2 + macr_densitym2 + 
                macj_densitym2 + nerj_densitym2 + ptej_densitym2 + lsetj_densitym2 + 
                tegula_densitym2 + cov_articulated_coralline + cov_crustose_coralline + 
                cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + 
                cov_dictyoneurum_spp + cov_dictyota_dictyopteris + cov_desmarestia_spp,
              data = year_one_data,
              CV = TRUE)

conf_mat <- table(Predicted = lda_cv$class, Actual = year_one_data$site_type)
acc_obs <- mean(lda_cv$class == year_one_data$site_type)

cat("\n---- Cross-Validation ----\n")
cat("Observed Accuracy:", round(acc_obs, 3), "\n")
cat("Confusion Matrix:\n")
print(conf_mat)

# ---- PERMUTATION TEST ----
acc_perm <- numeric(n_perm)

for (i in 1:n_perm) {
  shuffled <- year_one_data
  shuffled$site_type <- sample(shuffled$site_type)
  lda_perm <- lda(site_type ~ macro_stipe_density_ + densitymacstump + densityptecal + 
                    densitynerlue + densitylamset + densityeisarb + purple_urchin_densitym2 + 
                    purple_urchin_conceiledm2 + lamr_densitym2 + macr_densitym2 + 
                    macj_densitym2 + nerj_densitym2 + ptej_densitym2 + lsetj_densitym2 + 
                    tegula_densitym2 + cov_articulated_coralline + cov_crustose_coralline + 
                    cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + 
                    cov_dictyoneurum_spp + cov_dictyota_dictyopteris + cov_desmarestia_spp,
                  data = shuffled,
                  CV = TRUE)
  acc_perm[i] <- mean(lda_perm$class == shuffled$site_type)
}

p_value <- mean(acc_perm >= acc_obs)

cat("\n---- Permutation Test ----\n")
cat("Permutation p-value:", round(p_value, 4), "\n")

# ---- MANOVA ----
pred_scores <- predict(lda_model)$x
manova_res <- manova(pred_scores ~ year_one_data$site_type)
manova_summary <- summary(manova_res, test = "Wilks")

cat("\n---- MANOVA ----\n")
print(manova_summary)


# Get LDA scores
lda_scores <- as.data.frame(predict(lda_model)$x)
lda_scores$site_type <- year_one_data$site_type

# Plot LD1 vs LD2
ggplot(lda_scores, aes(x = LD1, y = LD2, color = site_type)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, linetype = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA: LD1 vs LD2",
    x = paste0("LD1 (", round(lda_model$svd[1]^2 / sum(lda_model$svd^2), 2) * 100, "%)"),
    y = paste0("LD2 (", round(lda_model$svd[2]^2 / sum(lda_model$svd^2), 2) * 100, "%)")
  )
```
###coefficients of LDA
```{r}
library(tidyr)

# Get coefficients into tidy format
coef_df <- as.data.frame(lda_model$scaling)
coef_df$variable <- rownames(coef_df)
coef_long <- pivot_longer(coef_df, cols = c(LD1, LD2), names_to = "LD", values_to = "coefficient")

# Plot
ggplot(coef_long, aes(x = reorder(variable, coefficient), y = coefficient, fill = LD)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "LDA Coefficients by Discriminant Function",
    x = "Variable",
    y = "Coefficient"
  ) +
  scale_fill_brewer(palette = "Set1")

```


##LDA with smallest replicable units 
###creating the new dataframe with merging the og dfs
```{r}

#QUESTION FOR JOSH: i think something is getting messed up here when i merge the dfs because quad data has transect but swath and macro data doesn't


merged_data_2 <- kelp_data %>%
  mutate(site = if_else(site == "RECO_04", "REC_04", site)) %>%
  filter(!(site == "REC_05" & site_type == "INCIP" & zone == "Shallow" & survey_date == as.Date("2025-06-09"))) %>%
  left_join(quad_transect_avgs, by = c("site", "site_type", "zone", "survey_date")) %>%
  left_join(macro_data, by = c("site","site_type", "zone", "survey_date"))


#creating a new column that just has years for future analysis  
merged_data_2$year <- as.numeric(format(merged_data_2$survey_date, "%Y"))


#normalizing 2024 data 

year_one_data_2 <- merged_data_2 %>% 
  filter(year == 2024) %>%
  filter() %>%
  dplyr::select(year, site, zone, site_type, macro_stipe_density_20m2, everything())

year_one_data_2[sapply(year_one_data_2, is.numeric)] <-
  lapply(year_one_data_2[sapply(year_one_data_2, is.numeric)],
         function(x) { x[is.na(x)] <- 0; x })

year_one_data_2 %>%
  group_by(site, site_type, zone, year) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")
```
###running the new LDA
```{r}
year_one_data_2$site_type <- as.factor(year_one_data_2$site_type)
  
# Identify columns that contain "20m2"
cols_to_scale <- grep("20m2", names(year_one_data_2), value = TRUE)

# Divide those columns by 20
year_one_data_2[cols_to_scale] <- lapply(year_one_data_2[cols_to_scale], function(x) x / 20)

# 3. Rename those columns by removing "20m2_"
names(year_one_data_2)[names(year_one_data_2) %in% cols_to_scale] <-
  gsub("20m2","", cols_to_scale)

lda_model_2 <- lda(site_type ~ macro_stipe_density_ + density_macstump + density_ptecal + density_nerlue + density_lamset + density_eisarb + purple_urchin_densitym2 + purple_urchin_conceiledm2 + lamr_densitym2 + macr_densitym2 + macj_densitym2 + nerj_densitym2 + ptej_densitym2 + lsetj_densitym2  + tegula_densitym2 + cov_articulated_coralline + cov_crustose_coralline + cov_encrusting_red + cov_fleshy_red + cov_stephanocystis + cov_dictyoneurum_spp + cov_dictyota_dictyopteris + cov_desmarestia_spp, data = year_one_data_2)

plot(lda_model_2)

lda_model_2

pred2 <- predict(lda_model_2)

pred_class2 <- pred2$class

table(Actual = year_one_data_2$site_type, Predicted = pred_class2)

results2 <- year_one_data_2 %>%
  mutate(predicted = pred_class2, 
         match = site_type == predicted)


mismatches2 <- results2 %>% filter(!match)



results2 <- results2 %>%
  dplyr::select(year, site, zone, site_type, predicted, match, everything())

results2

pred2 <- predict(lda_model_2)
table(Predicted = pred2$class, Actual = year_one_data_2$site_type)
mean(pred2$class == year_one_data_2$site_type)
```


#Bar Graphs
##kelps
```{r}
density_kelp_vars <- c(
  "macro_stipe_density_",
  "densityptecal",
  "densitynerlue", 
  "densitylamset", 
  "lamr_densitym2",
  "macr_densitym2",
  "macj_densitym2",
  "nerj_densitym2",
  "ptej_densitym2",
  "lsetj_densitym2"
)

# Define your variable labels by category
var_labels <- c(
  #MACRO
  macro_stipe_density_ = "M. pyrifera (adult)",
  macj_densitym2       = "M. pyrifera (juvenile)",
  macr_densitym2       = "M. pyrifera (recruit)",
  
  #NEREO
  densitynerlue        = "N. luetkeana (adult)", 
  nerj_densitym2       = "N. luetkeana (juvenile)",
  
  #PTECAL
  densityptecal        = "P. californicus (adult)",
  ptej_densitym2       = "P. californicus (juvenile)",
  
  #LSETJ
  densitylamset        = "L. setchellii (adult)",
  lsetj_densitym2      = "L. setchellii (juvenile)",
  
  lamr_densitym2       = "Laminaria (recruit)"
)

# Explicit order for plotting
plot_order <- c(
  "M. pyrifera (adult)",
    "M. pyrifera (juvenile)",
   "M. pyrifera (recruit)",
  

  "N. luetkeana (adult)",
    "N. luetkeana (juvenile)",
  
    "P. californicus (adult)",
    "P. californicus (juvenile)",
  
  "L. setchellii (adult)",
  "L. setchellii (juvenile)",
  
  "Laminaria (recruit)"
 
)

plot_kelp_data <- year_one_data %>%
  group_by(site, zone, site_type) %>%
  summarise(across(all_of(names(var_labels)), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(cols = all_of(names(var_labels)), names_to = "variable", values_to = "density") %>%
  group_by(variable, site_type) %>%
  summarise(
    mean_density = mean(density, na.rm = TRUE),
    se_density = sd(density, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    variable_label = factor(var_labels[variable], levels = plot_order) # explicit adult→juvenile→recruit order
  )

ggplot(plot_kelp_data, aes(x = variable_label, y = mean_density, fill = site_type)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.7, color = "black") +
  geom_errorbar(
    aes(ymin = mean_density - se_density, ymax = mean_density + se_density),
    position = position_dodge(width = 0.85),
    width = 0.2, size = 0.6
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Mean Kelp Density by Life Stage and Site Type",
    x = NULL,
    y = expression("Mean Density (individuals" ~ m^{-2}*")")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_blank(),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )


```


##urchins
```{r}
density_urchin_vars <- c(
  "purple_urchin_densitym2",
  "purple_urchin_conceiledm2"
)

urchin_var_labels <- c(
  purple_urchin_densitym2 = "Purple Urchin Density (m²)",
  purple_urchin_conceiledm2 = "Purple Urchin Conceiled Density (m²)")


plot_urchin_data <- year_one_data %>%
  group_by(site, zone, site_type) %>%
  summarise(across(all_of(names(urchin_var_labels)), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(cols = all_of(names(urchin_var_labels)), names_to = "variable", values_to = "density") %>%
  group_by(variable, site_type) %>%
  summarise(
    mean_density = mean(density, na.rm = TRUE),
    se_density = sd(density, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) 

plot_urchin_data <- plot_urchin_data %>%
  mutate(
    variable_label = factor(urchin_var_labels[variable], levels = urchin_var_labels)
  )

ggplot(plot_urchin_data, aes(x = variable_label, y = mean_density, fill = site_type)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.7, color = "black") +
  geom_errorbar(
    aes(ymin = mean_density - se_density, ymax = mean_density + se_density),
    position = position_dodge(width = 0.85),
    width = 0.2, size = 0.6
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Mean Urchin Density by Life Stage and Site Type",
    x = NULL,
    y = expression("Density")
  ) +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_blank(),
    legend.position = "top",
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 13),
    axis.text.y = element_text(size = 13),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```



#Tornado Plots
##BAR Deep
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "BAR"
selected_zone <- "Deep"

plot_data <- merged_data %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    purple_urchin_densitym2 = mean(purple_urchin_densitym2, na.rm = TRUE),
    purple_urchin_conceiledm2 = mean(purple_urchin_conceiledm2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

library(ggplot2)

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
BAR_deep <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Deep Barrens",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)


BAR_deep


```
##BAR Shallow
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "BAR"
selected_zone <- "Shallow"

plot_data <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
BAR_shallow <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Shallow Barrens",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)



BAR_shallow

```
##INCIP Deep
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "INCIP"
selected_zone <- "Deep"

plot_data <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
INCIP_deep <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Deep Incipients",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)

INCIP_deep

```
##INCIP Shallow
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "INCIP"
selected_zone <- "Shallow"

plot_data <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
INCIP_shallow <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Shallow Incipients",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)

INCIP_shallow

```
##FOR Deep
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "FOR"
selected_zone <- "Deep"

plot_data <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
FOR_deep <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Deep Forests",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)

FOR_deep

```
##FOR Shallow
```{r}
# Specify site_type and zone for this plot
selected_site_type <- "FOR"
selected_zone <- "Shallow"

plot_data <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025),
         site_type == selected_site_type,
         zone == selected_zone) %>%
  group_by(year) %>%  # group only by year since we're summarizing for this site_type+zone subset
  summarise(
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    ,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -year,
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value = ifelse(year == 2024, -value, value),  # 2024 values negative for left side
    year = as.factor(year)
  )

# Custom labels
var_labels <- c(
  total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)"
)

variable_order <- c(total_urchin_densitym2 = "Total Density of Urchins (m²)",
  total_urchin_foraging_densitym2 = "Density of Urchins Foraging (m²)",
  lamr = "Laminaria (recruit)",
  macr = "M. pyrifera (recruit)",
  macj = "M. pyrifera (juvenile)",
  nerj = "N. luetkeana (juvenile)",
  ptej = "P. californicus (juvenile)",
  lsetj = "L. setchellii (juvenile)",
  eisj = "E. arborea (juvenile)")


# Refined plot
FOR_shallow <- ggplot(plot_data, aes(x = reorder(variable, abs(value), levels = names(var_labels)), y = value, fill = year)) +
  geom_col(width = 0.7, color = "gray20") +
  coord_flip() +
  scale_y_continuous(labels = abs, expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_discrete(labels = var_labels) +
  scale_fill_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728")) +
  labs(
    title = "Shallow Forests",
    subtitle = "Mean Densities in 2024 vs 2025",
    x = NULL,
    y = "Mean Density (per m²)",
    fill = "Year"
  ) +
   theme_minimal(base_size = 14) +
theme(
  plot.title = element_text(hjust = 0.7, face = "bold", size = 16),
  plot.subtitle = element_text(hjust = 0.75, size = 13, margin = margin(b = 10)),
  plot.title.position = "plot",  # keeps title/subtitle together at top
  axis.text = element_text(color = "black"),
  legend.position = "top",
  panel.grid.major.y = element_blank()
)

FOR_shallow

```

#RDA for Kelp Recruits by Year, Site Type, and Zone

```{r}
library(tidyverse)
library(vegan)
library(ggrepel)  # for nice labels

# Step 1: Aggregate and prepare data (same as before)
agg_data <- quad_avgs %>%
  mutate(year = lubridate::year(survey_date)) %>%
  filter(year %in% c(2024, 2025)) %>%
  group_by(site_type, zone, year) %>%
  summarise(
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    eisj = mean(eisj, na.rm = TRUE),
    purple_urchin_densitym2 = mean(purple_urchin_densitym2, na.rm = TRUE),
    purple_urchin_active_foraging_densitym2 = mean(purple_urchin_active_foraging_densitym2, na.rm = TRUE),
    risk = mean(risk, na.rm = TRUE),
    relief = mean(relief, na.rm = TRUE),
    .groups = "drop"
  )

kelp_comm_agg <- agg_data %>%
  select(lamr, macr, macj, nerj, ptej, lsetj, eisj)

env_agg <- agg_data %>%
  select(purple_urchin_densitym2, purple_urchin_active_foraging_densitym2, risk, relief)

env_scaled <- as.data.frame(scale(env_agg))

# Step 2: Run RDA
rda_model <- rda(kelp_comm_agg ~ ., data = env_scaled)

# Step 3: Extract scores for plotting
sites <- as.data.frame(scores(rda_model, display = "sites"))
sites$site_type <- agg_data$site_type
sites$zone <- agg_data$zone
sites$year <- as.factor(agg_data$year)
sites$label <- paste(agg_data$site_type, agg_data$zone, agg_data$year, sep = " - ")

species <- as.data.frame(scores(rda_model, display = "species"))
species$species <- rownames(species)

biplot_env <- as.data.frame(scores(rda_model, display = "bp"))
biplot_env$variable <- rownames(biplot_env)

# Step 4: Plot with ggplot2
ggplot() +
  # Sites points
  geom_point(data = sites, aes(x = RDA1, y = RDA2, color = year, shape = zone), size = 4) +
  
  # Site labels
  ggrepel::geom_text_repel(data = sites, aes(x = RDA1, y = RDA2, label = label), size = 3, max.overlaps = 10) +
  
  # Species arrows and labels (scaled)
  geom_segment(data = species,
               aes(x = 0, y = 0, xend = RDA1*2, yend = RDA2*2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "darkgreen") +
  geom_text(data = species,
            aes(x = RDA1*2.2, y = RDA2*2.2, label = species),
            color = "darkgreen", size = 3, fontface = "italic") +
  
  # Environmental arrows and labels (scaled)
  geom_segment(data = biplot_env,
               aes(x = 0, y = 0, xend = RDA1*2, yend = RDA2*2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red") +
  geom_text(data = biplot_env,
            aes(x = RDA1*2.2, y = RDA2*2.2, label = variable),
            color = "red", size = 3, fontface = "bold") +
  
  # Axis labels and title
  labs(
    title = "RDA: Kelp Communities by Site Type, Zone, and Year",
    x = paste0("RDA1 (", round(summary(rda_model)$cont$importance[2,1]*100, 1), "% variance)"),
    y = paste0("RDA2 (", round(summary(rda_model)$cont$importance[2,2]*100, 1), "% variance)"),
    color = "Year",
    shape = "Zone"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  scale_color_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728"))

```

#RDA for Kelp Recruits by Year and Site Type

```{r}
library(tidyverse)
library(vegan)
library(ggrepel)
library(lubridate)

# Aggregate by site_type and year only (combining zones)
agg_data_site <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025)) %>%
  group_by(site_type, year) %>%
  summarise(
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    eisj = mean(eisj, na.rm = TRUE),
    purple_urchin_densitym2 = mean(purple_urchin_densitym2, na.rm = TRUE),
    purple_urchin_active_foraging_densitym2 = mean(purple_urchin_active_foraging_densitym2, na.rm = TRUE),
    risk = mean(risk, na.rm = TRUE),
    relief = mean(relief, na.rm = TRUE),
    .groups = "drop"
  )

# Kelp community matrix
kelp_comm_site <- agg_data_site %>%
  select(lamr, macr, macj, nerj, ptej, lsetj, eisj)

# Environmental predictors scaled
env_site <- agg_data_site %>%
  select(purple_urchin_densitym2, purple_urchin_active_foraging_densitym2, risk, relief)

env_scaled_site <- as.data.frame(scale(env_site))

# RDA model
rda_model_site <- rda(kelp_comm_site ~ ., data = env_scaled_site)

# Extract scores for plotting
sites <- as.data.frame(scores(rda_model_site, display = "sites"))
sites$site_type <- agg_data_site$site_type
sites$year <- as.factor(agg_data_site$year)
sites$label <- paste(agg_data_site$site_type, agg_data_site$year, sep = " - ")

species <- as.data.frame(scores(rda_model_site, display = "species"))
species$species <- rownames(species)

biplot_env <- as.data.frame(scores(rda_model_site, display = "bp"))
biplot_env$variable <- rownames(biplot_env)

# Plot with ggplot2
ggplot() +
  geom_point(data = sites, aes(x = RDA1, y = RDA2, color = year), size = 5) +
  ggrepel::geom_text_repel(data = sites, aes(x = RDA1, y = RDA2, label = label), size = 4) +
  
  geom_segment(data = species,
               aes(x = 0, y = 0, xend = RDA1*2, yend = RDA2*2),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "darkgreen") +
  geom_text(data = species,
            aes(x = RDA1*2.2, y = RDA2*2.2, label = species),
            color = "darkgreen", size = 4, fontface = "italic") +
  
  geom_segment(data = biplot_env,
               aes(x = 0, y = 0, xend = RDA1*2, yend = RDA2*2),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "red") +
  geom_text(data = biplot_env,
            aes(x = RDA1*2.2, y = RDA2*2.2, label = variable),
            color = "red", size = 4, fontface = "bold") +
  
  labs(
    title = "RDA: Kelp Communities by Site Type and Year",
    x = paste0("RDA1 (", round(summary(rda_model_site)$cont$importance[2,1]*100, 1), "% variance)"),
    y = paste0("RDA2 (", round(summary(rda_model_site)$cont$importance[2,2]*100, 1), "% variance)"),
    color = "Year"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  scale_color_manual(values = c("2024" = "#1f77b4", "2025" = "#d62728"))
  
```
```{r}
# Aggregate & filter data
agg_data_site <- quad_avgs %>%
  mutate(year = year(survey_date)) %>%
  filter(year %in% c(2024, 2025)) %>%
  filter(!site %in% c("REC_06", "REC_07")) %>%   # <--- remove these sites
  group_by(site, site_type, zone, year) %>%
  summarise(
    lamr = mean(lamr, na.rm = TRUE),
    macr = mean(macr, na.rm = TRUE),
    macj = mean(macj, na.rm = TRUE),
    nerj = mean(nerj, na.rm = TRUE),
    ptej = mean(ptej, na.rm = TRUE),
    lsetj = mean(lsetj, na.rm = TRUE),
    eisj = mean(eisj, na.rm = TRUE),
    total_urchin_densitym2 = mean(total_urchin_densitym2, na.rm = TRUE),
    total_urchin_foraging_densitym2 = mean(total_urchin_foraging_densitym2, na.rm = TRUE),
    risk = mean(risk, na.rm = TRUE),
    relief = mean(relief, na.rm = TRUE),
    .groups = "drop"
  )

# Prepare data
kelp_comm <- agg_data_site %>% select(lamr, macr, macj, nerj, ptej, lsetj, eisj)
kelp_comm_hel <- decostand(kelp_comm, method = "hellinger")
rda_model <- rda(kelp_comm_hel ~ ., data = env_vars)

env_vars <- agg_data_site %>%
  select(total_urchin_densitym2, total_urchin_foraging_densitym2, risk, relief) %>%
  scale() %>%
  as.data.frame()

# Run RDA

# Extract scores
site_scores <- as.data.frame(scores(rda_model, display = "sites"))
site_scores <- bind_cols(site_scores, agg_data_site %>% select(site, site_type, zone, year)) %>%
  mutate(label = paste(site, zone), year = as.factor(year))

species_scores <- as.data.frame(scores(rda_model, display = "species", scaling = 2))
env_scores <- as.data.frame(scores(rda_model, display = "bp", scaling = 2))

# Colorblind-friendly palette (Okabe-Ito)
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot() +
  # Ellipses (lighter fill for clarity)
  stat_ellipse(data = site_scores, aes(x = RDA1, y = RDA2, group = interaction(site_type, year), color = site_type, fill = site_type),
               linetype = "dashed", alpha = 0.15, size = 1, geom = "polygon", show.legend = FALSE) +
  # Site points
  geom_point(data = site_scores, aes(x = RDA1, y = RDA2, color = site_type, shape = year),
             size = 5, stroke = 1, alpha = 0.85) +
  # Site labels (optional: comment out if too busy)
  #geom_text(data = site_scores, aes(x = RDA1, y = RDA2, label = label),
            #size = 3, vjust = -1, hjust = 0.5, color = "gray20", fontface = "bold") +
  # Species vectors (italicized labels)
  geom_segment(data = species_scores, aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.3, "cm")), color = "#009E73", linewidth = 1) +
  geom_text(data = species_scores, aes(x = RDA1, y = RDA2, label = rownames(species_scores)),
            color = "#009E73", fontface = "italic", size = 4, vjust = 1.5) +
  # Environmental arrows (bold labels)
  geom_segment(data = env_scores, aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.3, "cm")), color = "#D55E00", linewidth = 1) +
  geom_text(data = env_scores, aes(x = RDA1, y = RDA2, label = rownames(env_scores)),
            color = "#D55E00", size = 4, vjust = 1.2, fontface = "bold") +
  # Labels and theme
  labs(
    title = "RDA of Kelp Community Composition (2024–2025)",
    subtitle = "Environmental predictors: Urchin density, foraging, risk, and relief",
    x = "RDA1 (constrained variance)",
    y = "RDA2",
    color = "Site Type",
    shape = "Year"
  ) +
  scale_color_manual(values = cb_palette) +
  scale_fill_manual(values = cb_palette) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "top",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 14, margin = margin(b = 15)),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(color = "black"),
    panel.grid.minor = element_blank()
  )

site_scores %>% arrange(desc(RDA1)) %>% head()
site_scores %>% arrange(RDA1) %>% head()


```



#Swath Data Cleaning
```{r}

#just selecting the columns i'm interested in
kelp_clean <- kelp_raw %>%
  dplyr::select(site, site_type, zone, date, species, stipe_counts_macrocystis_only, count, transect, subsample_meter) %>% 
  group_by(site, site_type, zone, date, species, stipe_counts_macrocystis_only, count, transect, subsample_meter) 
    
#fixing the problems with the subsampling. subsampling meter cannot be 28.6 so i think the diver forgot to do the math underwater

kelp_clean[8, "subsample_meter"] <- 8.6
kelp_clean[72, "subsample_meter"] <- 5
kelp_clean[59, "subsample_meter"] <- 3.9
kelp_clean[211, "subsample_meter"] <- 9.8
kelp_clean[73, "subsample_meter"] <- 9.0
kelp_clean[34, "subsample_meter"] <- 8.4
kelp_clean[53, "subsample_meter"] <- 6.3

#calculating subsample fraction (subsample meter/ 10m)
kelp_clean <- kelp_clean %>%
  mutate(sampling_fraction = subsample_meter / 10)
  
#this is just changing the NAs on sampling fraction to 1 so that counts which werent subsampled wont be void after these calculations
kelp_clean$sampling_fraction[is.na(kelp_clean$sampling_fraction)] <- 1

#this is the adjusted counts after factoring in the subsample meter
kelp_clean <- kelp_clean %>%
  mutate(adjusted_count = count / sampling_fraction)   

#removing purps on kelp because i only want to focus on plants for this analysis
kelp_clean <- kelp_clean %>%
  filter(species != "purps_on_kelp")
  
#changing the time scale from specific date to just year
kelp_cleaner <- kelp_clean %>%
  mutate(date = ymd(date),
         year = year(date)) %>%
  group_by(site, site_type, zone, year, species, stipe_counts_macrocystis_only, adjusted_count)

#just sorting by site_type, zone, and year
kelp_species <- kelp_cleaner %>%
  ungroup() %>%   # REMOVE any existing grouping
  mutate(year = year(ymd(date))) %>%
  select(site_type, zone, year, species, adjusted_count) %>%
  arrange(site_type, zone, year) %>%
  group_by(site_type, zone, year, species, adjusted_count)
#removed macro stipe counts just for sake of only having one column of count values

#now combining counts by site type, zone, year, and species
kelp_summary <- kelp_species %>%
  ungroup() %>%   # just to be safe, remove any grouping
  group_by(site_type, zone, year, species) %>%
  summarise(total_adjusted_count = sum(adjusted_count, na.rm = TRUE), .groups = "drop")

view(kelp_summary)     
```

#plotting constrained ordination analysis
```{r}
kelp_clean_ungrouped <- kelp_cleaner %>% 
  ungroup() # remove grouping metadata

kelp_wide <- kelp_clean_ungrouped %>%
  group_by(site_type, zone, year, species) %>%
  summarise(adjusted_count = sum(adjusted_count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = species, values_from = adjusted_count, values_fill = 0)

kelp_wide

species_matrix <- kelp_wide %>%
  select(-site_type, -zone, -year)  # all species columns

explanatory_matrix <- kelp_wide %>%
  select(site_type)  # or add zone/year if needed

str(kelp_wide %>% select(site_type, zone, year))
explanatory_vars <- kelp_wide %>%
  mutate(
    site_type = factor(site_type),
    zone = factor(zone),
    year = as.numeric(year)
  ) %>%
  select(site_type, zone, year)
species_matrix <- kelp_wide %>% select(-site_type, -zone, -year)

rda_model <- vegan::rda(species_matrix ~ site_type + zone + year, data = explanatory_vars)
summary(rda_model)
plot(rda_model)



# Convert site_type and zone to factors, year to numeric
kelp_wide_clean <- kelp_wide %>%
  mutate(
    site_type = factor(site_type),
    zone = factor(zone),
    year = as.numeric(year)
  )

# Split explanatory and response matrices
explanatory_vars <- kelp_wide_clean %>% select(site_type, zone, year)
species_matrix <- kelp_wide_clean %>% select(-site_type, -zone, -year)

sum(is.na(explanatory_vars))   # should be 0
sum(is.na(species_matrix))     # MUST be 0, or RDA will fail

species_matrix[is.na(species_matrix)] <- 0


rda_model <- rda(species_matrix ~ site_type + zone + year, data = explanatory_vars)
summary(rda_model)
plot(rda_model)


colSums(is.na(kelp_wide_clean %>% select(site_type, zone, year)))

kelp_filtered <- kelp_wide_clean %>% 
  filter(!is.na(site_type))

# Split into explanatory and species matrices
explanatory_vars <- kelp_filtered %>% select(site_type, zone, year)
species_matrix <- kelp_filtered %>% select(-site_type, -zone, -year)

# Confirm no NAs now
sum(is.na(explanatory_vars))     # should be 0
sum(is.na(species_matrix))       # should be 0

# Run RDA
library(vegan)
rda_model <- rda(species_matrix ~ site_type + zone + year, data = explanatory_vars)
summary(rda_model)
plot(rda_model, scaling = 2)

```




```{r}
# --- Run RDA ---
rda_result <- rda(species_matrix ~ site_type * year + zone, data = explanatory_vars)

# --- Calculate % variance explained for axes ---
eigvals <- rda_result$CCA$eig
var_exp <- round(100 * eigvals / sum(eigvals), 1)
rda.axis.long <- data.frame(
  axis = c("RDA1", "RDA2"),
  label = paste0(c("RDA1", "RDA2"), " (", var_exp[1:2], "%)")
)

# --- Extract site scores and add grouping variables ---
site_scores <- as.data.frame(scores(rda_result, display = "sites"))
site_scores$site_type <- explanatory_vars$site_type
site_scores$year <- factor(explanatory_vars$year)

# --- Extract species scores and filter top loadings ---
species_scores <- as.data.frame(scores(rda_result, display = "species"))
loading_threshold <- 0.4
top_species_scores <- species_scores %>%
  filter(abs(RDA1) > loading_threshold | abs(RDA2) > loading_threshold) %>%
  mutate(species = rownames(.))

# --- Plot ---
rda_plot <- ggplot() +
  geom_vline(xintercept = 0, color = "grey70", linetype = 2) +
  geom_hline(yintercept = 0, color = "grey70", linetype = 2) +
  xlab(rda.axis.long$label[1]) +
  ylab(rda.axis.long$label[2]) +
  scale_x_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +
  scale_y_continuous(sec.axis = dup_axis(labels = NULL, name = NULL)) +

  # Points: shape = site_type, color = year, white fill and black outline
  geom_point(data = site_scores,
             aes(x = RDA1, y = RDA2, shape = site_type, color = year),
             size = 4, fill = "white", stroke = 1) +

  # Species arrows
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.1, "inches")),
               size = 0.6, color = "blue") +

  # Species labels
  geom_text_repel(data = top_species_scores,
                  aes(x = RDA1, y = RDA2, label = species),
                  size = 4, color = "blue", segment.color = "grey70") +

  scale_shape_manual(name = "Site Type",
                     values = c("BAR" = 21, "FOR" = 22, "INCIP" = 24)) +
  scale_color_manual(name = "Year", 
                     values = c("2024" = "blue", "2025" = "red", "2026" = "darkgray")) +

  theme_minimal(base_size = 14) +
  coord_fixed() +
  theme(legend.position = "right")

print(rda_plot)
```